---
title: "Hayman Fire Recovery"
author: "Anna Marshall"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
---
# Overview
The Hayman Fire was a forest fire started on June 8, 2002, 35 miles northwest of Colorado Springs, Colorado. Below is a graphical analysis looking at some pre and post-fire changes. 

```{r setup, results = FALSE, results = FALSE, echo=FALSE, fig.show='hide',warning=FALSE,message=FALSE}
library(tidyverse)
library(tidyr)
library(ggthemes)
library(lubridate)
library(dplyr)

# Now that we have learned how to munge (manipulate) data
# and plot it, we will work on using these skills in new ways

knitr::opts_knit$set(root.dir='..')

```

```{r results = FALSE, echo=FALSE, fig.show='hide',warning=FALSE,message=FALSE}
#Reading in the data. 
ndvi <- read_csv('data/hayman_ndvi.csv') %>%
  rename(burned=2,unburned=3) %>%
  filter(!is.na(burned),
         !is.na(unburned))

# Converting from wide to long data
ndvi_long <- gather(ndvi,
                    key='site',
                    value='NDVI',
                    -DateTime)

# Plotting all the data
ggplot(ndvi_long,aes(x=DateTime,y=NDVI,color=site))+
  geom_point(shape=1) +
  geom_line() +
  theme_few() + 
  scale_color_few() +
  theme(legend.position=c(0.3,0.3))

# Summarizing the data by year
ndvi_annual <- ndvi_long %>%
  mutate(year=year(DateTime)) %>%
  mutate(month=month(DateTime)) %>%
  filter(month %in% c(5,6,7,8,9)) %>%
  group_by(site,year) %>%
  summarize(mean_NDVI=mean(NDVI))

#Here making an annual plot
ggplot(ndvi_annual,aes(x=year,y=mean_NDVI,color=site)) +
  geom_point(shape=1) + 
  geom_line() +
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.3,0.3))


# Plotting seasonal variation by summarizing over month instead of year
ndvi_month <- ndvi_long %>%
  mutate(year=year(DateTime)) %>%
  mutate(month=month(DateTime)) %>%
  group_by(site,month) %>%
  summarize(mean_NDVI=mean(NDVI))

# Same plot as above but with month on x-axis
ggplot(ndvi_month,aes(x=month,y=mean_NDVI,color=site)) +
  geom_point(shape=1) + 
  geom_line() +
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.6,0.2))


## Adding another groupd called treatment (pre or post-burn)
ndvi_month_pre_post <- ndvi_long %>% 
  mutate(year = year(DateTime),
         month = month(DateTime),
         treatment = cut(year,breaks=c(0,2003,2020),
                         labels=c('pre-burn','post-burn'))) %>%
  group_by(month,site,treatment) %>%
  summarize(mean_ndvi = mean(NDVI))

# Plot that splits out data by burned and unburned (facet_wrap(~site))            
ggplot(ndvi_month_pre_post,aes(x=month,y=mean_ndvi,color=treatment)) +
  geom_point(shape=1) + 
  geom_line() +
  theme_few() + 
  scale_color_few() + 
  theme(legend.position=c(0.6,0.2)) + 
  facet_wrap(~site)

```

```{r echo=FALSE, fig.show='hide',results = FALSE, warning=FALSE,message=FALSE}
#Reading in files
files <- list.files('data',full.names=T)

#Read in individual data files
ndmi <- read_csv(files[1]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndmi')


ndsi <- read_csv(files[2]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndsi')

ndvi <- read_csv(files[3])%>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndvi')

## join all datasets together and rename columns
full_wide <- inner_join(ndmi %>% dplyr::select(-data),
                        ndvi %>% dplyr::select(-data),
                        by='DateTime') %>%
  inner_join(ndsi %>% dplyr::select(-data),by='DateTime') %>%
  rename(burned_ndmi = 2,unburned_ndmi=3,
         burned_ndvi=4, unburned_ndvi=5,
         burned_ndsi=6, unburned_ndsi=7) %>%
  filter_if(is.numeric,all_vars(!is.na(.))) %>% 
  mutate(month=month(DateTime),
         year=year(DateTime))


## Plot of burned plots of moisture (ndmi) vs greenness (ndvi)
full_wide %>%
  filter(!month %in% c(11,12,1,2,3,4,5)) %>%
ggplot(.,aes(x=burned_ndmi,y=burned_ndvi,color=month)) + 
  geom_point()


# Stack as a tidy dataset
full_long <- rbind(ndvi,ndmi,ndsi) %>%
  gather(key='site',value='value',-DateTime,-data) %>%
  filter(!is.na(value))



# Plot all three different types
ggplot(full_long,aes(x=DateTime,y=value,color=site)) + 
  geom_line() + 
  facet_wrap(~data)+
  theme_few() + 
  scale_color_few() 

```
## Q1. Correlation between NDVI and NDMI
The NDVI is correlated with the NDMI. A higher NDMI should result in a higher NDVI. In other words, more moisture will result in a higher amount of greeness. More precipitation and wetter soils means vegetation can grow more effectively. However, we see a different relationship post-burn. 
```{r warning=FALSE,message=FALSE}
##convert full_long dataset to wide format
data_wide<-spread(full_long,data,value)%>%
  mutate(month=month(DateTime),
         year = year(DateTime),
         treatment = cut(year,breaks=c(0,2003,2020),
                         labels=c('pre-burn','post-burn')))
        

## Plot of moisture (ndmi) vs greenness (ndvi) for burned area
data_wide %>%
  filter(!month %in% c(11,12,1,2,3,4,5)) %>%
ggplot(.,aes(x=ndmi,y=ndvi,color=month)) + 
  geom_point()+
  facet_wrap(~site)+
  theme_few()

```

## Q2. Correlation between average NDSI (January-April) and NDVI (June-August)
The average NDSI during the winter is correlated to the NDVI during the summer. More snow means more NDVI in the summer. 
```{r warning=FALSE,message=FALSE}
## Plot of snow (ndsi) vs greenness (ndvi) 
data_wide %>%
  filter(!month %in% c(5,9,10,11,12)) %>%
ggplot(.,aes(x=ndvi,y=ndsi,color=month)) + 
  geom_point()+
  theme_few()
```



## Q3. Comparing snow effect between burned vs. unburned and pre- vs. post-burn.

```{r warning=FALSE,message=FALSE}
## Plot of snow (ndsi) vs greenness (ndvi) seperated by burned vs. unburned
data_wide %>%
  filter(!month %in% c(5,9,10,11,12)) %>%
ggplot(.,aes(x=ndvi,y=ndsi,color=month)) + 
  geom_point()+
  facet_wrap(~site)+
  theme_few()

## Plot of snow (ndsi) vs greenness (ndvi) seperated by pre- vs. post-burn
##Using 1984-2001 as pre-burn and 2002-2019 as post-burn
data_wide %>%
  filter(!month %in% c(5,9,10,11,12)) %>%
ggplot(.,aes(x=ndvi,y=ndsi,color=month)) + 
  geom_point()+
  facet_wrap(~treatment)+
  theme_few()


```

## Q4. What month is the greenest month on average? 
July appears to be the greenest month on average in both the burned and unburned zones. 
```{r warning=FALSE,message=FALSE}
# Plotting seasonal variation by summarizing over month instead of year
ndvi <- read_csv('data/hayman_ndvi.csv') %>%
  rename(burned=2,unburned=3) %>%
  filter(!is.na(burned),
         !is.na(unburned))

ndvi_long <- gather(ndvi,
                    key='site',
                    value='NDVI',
                    -DateTime)

ndvi_month <- ndvi_long %>%
  mutate(year=year(DateTime)) %>%
  mutate(month=month(DateTime)) %>%
  group_by(site,month) %>%
  summarize(mean_NDVI=mean(NDVI))

ggplot(ndvi_month,aes(x=month,y=mean_NDVI,color=site)) +
  geom_point(shape=1) + 
  geom_line() +
  theme_few() + 
  scale_color_few() + 
  scale_x_continuous(breaks=seq(1,12,1))+
  theme(legend.position=c(0.6,0.2))
```


## Q5. What month is the snowiest on average?
In the unburned zone, February has the highest NDSI, while NDSI in the burned zone suggests a snowier January. 
```{r warning=FALSE,message=FALSE}
#Reading in the data. 
ndsi <- read_csv('data/hayman_ndsi.csv') %>%
  rename(burned=2,unburned=3) %>%
  filter(!is.na(burned),
         !is.na(unburned))

# Converting from wide to long data
ndsi_long <- gather(ndsi,
                    key='site',
                    value='NDSI',
                    -DateTime)
# Plotting seasonal variation by summarizing over month instead of year
ndsi_month <- ndsi_long %>%
  mutate(year=year(DateTime)) %>%
  mutate(month=month(DateTime)) %>%
  group_by(site,month) %>%
  summarize(mean_NDSI=mean(NDSI))

# Same plot as above but with month on x-axis
ggplot(ndsi_month,aes(x=month,y=mean_NDSI,color=site)) +
  geom_point(shape=1) + 
  geom_line() +
  theme_few() + 
  scale_color_few() + 
  scale_x_continuous(breaks=seq(1,12,1))+
  theme(legend.position=c(0.6,0.2))
```



## Bonus Question: Redo all problems with `spread` and `gather` using modern tidyverse syntax. 
Modern tidyverse syntax uses pivot_wider() and pivot_longer() to reshape data between wide and long formats. Check out the updated code below that utilizes this modern tidyverse syntax. 
```{r warning=FALSE,message=FALSE, results = FALSE}
#convert long data to wide
wide_pivot<-pivot_wider(full_long,names_from = "site", values_from = "value")%>%
  mutate(month=month(DateTime),
         year = year(DateTime),
         treatment = cut(year,breaks=c(0,2003,2020),
                         labels=c('pre-burn','post-burn')))
#convert wide data to long
full_wide%>% 
  pivot_longer(burned_ndmi:unburned_ndsi,
              names_to="data", 
              values_to='value',
              values_drop_na = TRUE)
```

## Bonus Question: Use Climage Engine to pull the same data for the assignment, but updated with 2020/2021 data.
Data has been updated in all graphs to include 2020/2021 data for NDVI, NDSI, and NDMI. 



